---
title: "Statistical Inference"
author: "Nguyen Toan"
date: "October 26th, 2014"
output: html_document
---
  
# Part 2: Basic inferential data analysis
  
## 1. Load the ToothGrowth data and perform some basic exploratory data analysis     
  
```{r}
# load the data
library(datasets)
data(ToothGrowth)

# some basic exploratory data analyses
head(ToothGrowth)
```

By using `?ToothGrowth`, we can get the explanations on the data.
```
A data frame with 60 observations on 3 variables.
  [,1]   len	 numeric	 Tooth length
  [,2]	 supp	 factor	 Supplement type (VC or OJ).
  [,3]	 dose	 numeric	 Dose in milligrams.
```

## 2. Provide a basic summary of the data

```{r}
summary(ToothGrowth)
```  

## 3. Use confidence intervals and hypothesis tests to compare tooth growth by supp and dose.  

We create a linear regression model with `len` explained by `dose` and `supp` and calculate the 95% confidence intervals for the coefficients.
  
```{r}
fit <- lm(len ~ dose + supp, data=ToothGrowth)
confint(fit, level=0.95)
```

The result means that 95% of the time which we collect a different set of data and estimate parameters of the linear regression model, the coefficient estimations will vary in these confidence intervals.

```{r}
summary(fit)
```

Here we consider the null hypothesis, which assumes that the coefficients in the linear regression model are zeros.
From the summary of the model, we see that all _p_-values are less than 0.05, which means the null hypothesis is rejected with the 5% significance level. In other words,  each variable significantly explains the variability in tooth length.

For example, the coefficient of `dose` is `r fit$coefficients[[2]]`, which means that increasing the dose 1 mg (while fixing `supp`) would increase the tooth length `r fit$coefficients[[2]]` units.
